Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-02-22 paperDistance/paper.tex
acronym trouble

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}

@@

Here we build and use just such a data-driven model.
In particular, we use all of the \gaia\ \tgas\ data, matched to the \tmass\
photometry, to make a model of the noise-free (or low-noise)
color--magnitude diagram (CMD) of stars.
We use
an Empirical-Bayes approach known as Extreme Deconvolution (DWH CITE).
This method deconvolves heteroskedastic data to derive an estimate
of the noise-free or low-noise distribution that \emph{would have}
been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (DWH CITE BOTH) and to perform quasar target selection
in \project{Sloan Digital Sky Survey} data (DWH CITE BOTH).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the CMD to a space in which
the noise is approximately Gaussian.

The novel use of XD here is to treat its output---the deconvolved CMD---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the XD model to de-noise the \gaia\ parallax
measurements.
Technically, since XD produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the XD approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to
a related project (CITE BORIS), in which we use a
somewhat less flexible model for the CMD, but in which we take a much more principled
approach to the inference.

@@

Dust corrections are non trivial for stars with poor signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giants, which tend to have the lowest signal to noise, and can therefore have dust corrections $\sim 1-2$ magnitudes. To apply a dust correction to the 2MASS photometry, we first generate our prior using XD on the observed, attenuated photometry. Using this raw prior we infer the distances to all the stars in our prior more precisely, increasing our signal to noise for each star and therefore minimize severe distances and their associated severe dust corrections. With more precise distance posteriors, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the Green et al 3D dust map at that distance. For each distance we take the median of the samples of the dust model. We apply each dust correction to our 2MASS photometry given by Equation \ref{eq:dust} where $K_{\lambda}$ = [3.626, 2.742, 3.303, 2.285, 1.698] for bands [B, V, g, r, i] respectively, taken from table (REF TABLE SFD), and $E(B-V)$ is the output of the Green et al dust models. We then regenerate our prior using the dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads again to a lighter deconvolution. Should I quote some statistics about the dust corrections? mostly small? maybe difference before after correcting for distance?

@@

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ data without any use of physical stellar
models, nor models of the Milky Way.
We obtained these by building a data-driven model of the
color--magnitude diagram (CMD) of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian interences.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown thatm, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in stellar
clusters are much more clustered than likelihood-based distance
estimates.

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's distance by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution makes use of a reasonable noise model and an
assumption of stationarity to shrink the parallax uncertainties, and
do so dramatically for the stars measured at lower signal-to-noise.
Because the model we use (Extreme Deconvolution or XD) accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the color--magnitude diagram (CMD)---using
all the data, not just the data with the highest signal-to-noise or
largest parallaxes.
That is, the model for the CMD we have built is representative for
the \tgas\ Catalog.
Inasmuch as that is true, we expect any (sensible) distance estimates
we generate from our distance posterior pdfs to be weakly biased.

@@

Number of gaussians in XD, and the w parameter. That we set them heuristically.

>>>>

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

Here we build and use just such a data-driven model.
In particular, we use all of the \gaia\ \tgas\ data, matched to the \tmass\
photometry, to make a model of the noise-free (or low-noise)
color--magnitude diagram (\cmd) of stars.
We use
an Empirical-Bayes approach known as Extreme Deconvolution (DWH CITE).
This method deconvolves heteroskedastic data to derive an estimate
of the noise-free or low-noise distribution that \emph{would have}
been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (DWH CITE BOTH) and to perform quasar target selection
in \project{Sloan Digital Sky Survey} data (DWH CITE BOTH).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

The novel use of \xd\ here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to
a related project (CITE BORIS), in which we use a
somewhat less flexible model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

Dust corrections are non trivial for stars with poor signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giants, which tend to have the lowest signal to noise, and can therefore have dust corrections $\sim 1-2$ magnitudes. To apply a dust correction to the 2MASS photometry, we first generate our prior using \xd\ on the observed, attenuated photometry. Using this raw prior we infer the distances to all the stars in our prior more precisely, increasing our signal to noise for each star and therefore minimize severe distances and their associated severe dust corrections. With more precise distance posteriors, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the Green et al 3D dust map at that distance. For each distance we take the median of the samples of the dust model. We apply each dust correction to our 2MASS photometry given by Equation \ref{eq:dust} where $K_{\lambda}$ = [3.626, 2.742, 3.303, 2.285, 1.698] for bands [B, V, g, r, i] respectively, taken from table (REF TABLE SFD), and $E(B-V)$ is the output of the Green et al dust models. We then regenerate our prior using the dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads again to a lighter deconvolution. Should I quote some statistics about the dust corrections? mostly small? maybe difference before after correcting for distance?

@@

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ data without any use of physical stellar
models, nor models of the Milky Way.
We obtained these by building a data-driven model of the
color--magnitude diagram (\cmd) of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian interences.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown thatm, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in stellar
clusters are much more clustered than likelihood-based distance
estimates.

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's distance by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution makes use of a reasonable noise model and an
assumption of stationarity to shrink the parallax uncertainties, and
do so dramatically for the stars measured at lower signal-to-noise.
Because the method we use (Extreme Deconvolution or \xd, which is
an Empirical Bayesian maximum-marginalized-likelihood estimator) accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the \cmd---using
all the data, not just the data with the highest signal-to-noise or
largest parallaxes.
That is, the model for the \cmd\ we have built is representative for
the \tgas\ Catalog.
Inasmuch as that is true, we expect any (sensible) distance estimates
we generate from our distance posterior pdfs to be weakly biased.

@@

Number of gaussians in \xd, and the w parameter. That we set them heuristically.
